---
title: "Sentiment Analysis of Student Reflections"
author: "Oskar Diyali"
output: html_document
---

## Objective

This analysis explores the sentiment expressed in 20 student reflection responses to a common question:  
**"What did you learn or experience from this task?"** (for now)

Two approaches are used for sentiment analysis:
- **AFINN Lexicon** via `tidytext` to score sentiment at the sentence level
- **NRC Emotion Lexicon** via `syuzhet` to classify emotional categories

---

## 1. Load Required Libraries

```{r 1}

library(tidyverse)
library(tidytext)
library(syuzhet)
library(readr)
library(ggplot2)

```

## 2. Load the Data
```{r 2}

# Load student reflections including the question column
reflections <- read_csv("data/student_reflections_with_question_long.csv")


```


## 3. Sentence Tokenization
```{r 3}

# Split each reflection into individual sentences
sentences <- reflections %>%
  unnest_tokens(sentence, reflection, token = "sentences")


```


## 4. Word Tokenization
```{r 4}

# Assign a unique sentence ID and tokenize each sentence into words
words <- sentences %>%
  mutate(sentence_id = row_number()) %>%
  unnest_tokens(word, sentence)


```


## 5. AFINN Sentiment Analysis
```{r 5}

# Load AFINN lexicon
afinn <- get_sentiments("afinn")

# Join with tokenized words to assign sentiment values
scored_words <- words %>%
  inner_join(afinn, by = "word")



```

## 6. Sum Sentiment Per Sentence

```{r 6}

# Calculate total sentiment score per sentence
sentence_scores <- scored_words %>%
  group_by(sentence_id) %>%
  summarise(sentiment = sum(value), .groups = "drop") %>%
  left_join(words %>% distinct(sentence_id, student_id), by = "sentence_id") %>%
  left_join(reflections, by = "student_id") %>%
  select(student_id, question, sentence_id, sentiment, reflection)

sentence_scores

```


## 7. Top Positive Sentences (AFINN)
```{r 7}

# Identify the five most positive-scoring sentences
top_sentences <- sentence_scores %>%
  arrange(desc(sentiment)) %>%
  distinct(sentence_id, .keep_all = TRUE) %>%
  slice_head(n = 5)

top_sentences


```

## 8. Histogram of AFINN Sentiment Scores
```{r 8}

# Histogram for sentiment scores
ggplot(sentence_scores, aes(x = sentiment)) +
  geom_histogram(
    binwidth = 1,
    fill = "#0072B2",
    color = "white",
    boundary = 0,
    closed = "left"
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(), 
    expand = expansion(mult = c(0, 0))  
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +  
  labs(
    title = "Distribution of AFINN Sentiment Scores per Sentence",
    x = "AFINN Sentiment Score",
    y = "Number of Sentences"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black", linewidth = 0.5),
    panel.grid = element_blank()
  )




```

## 9. NRC Emotion Analysis using 'syuzhet'

```{r 9}

# Apply NRC emotion lexicon to full reflections
nrc_emotions <- get_nrc_sentiment(reflections$reflection)

# Combine emotions with student metadata
nrc_df <- bind_cols(reflections %>% select(student_id, question), nrc_emotions)

head(nrc_df)




```


## 10. Emotion Summary Plot (NRC)

```{r 10}

ggplot(nrc_summary, aes(x = reorder(emotion, count), y = count)) +
  geom_col(fill = "#D55E00", width = 0.8) +
  geom_text(aes(label = count), hjust = -0.1, size = 4.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +  
  labs(
    title = "Frequency of Emotion Words in Student Reflections (NRC)",
    x = "Emotion Category",
    y = "Word Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    plot.margin = margin(10, 30, 10, 10)  
  )




```