---
title: "Sentiment Analysis of Student Reflections"
author: "Oskar Diyali"
output: html_document
---

## 📌 Objective

This analysis evaluates the sentiment of 20 student (for now) reflection responses using two approaches:
- **AFINN Lexicon** via `tidytext` for sentence-level sentiment scoring
- **NRC Emotion Lexicon** via `syuzhet` for emotion classification  
It also analyzes sentiment in the context of specific reflection questions.

---

## 1. Load Libraries

```{r}
library(tidyverse)
library(tidytext)
library(syuzhet)
library(readr)
library(ggplot2)
```

## 2. Load the Data
```{r}
# Load reflection responses that include reflection question
reflections <- read_csv("student_reflections_with_question.csv")

# Preview the dataset
head(reflections)


```


## 3. Sentence Tokenization
```{r}

# Break reflections into individual sentences
sentences <- reflections %>%
  unnest_tokens(sentence, reflection, token = "sentences")


```


## 4. Word Tokenization
```{r}

# Assign unique sentence IDs and tokenize into words
words <- sentences %>%
  mutate(sentence_id = row_number()) %>%
  unnest_tokens(word, sentence)


```


## 5. AFINN Sentiment Analysis
```{r}

# Load AFINN lexicon
afinn <- get_sentiments("afinn")

# Join with tokenized words
scored_words <- words %>%
  inner_join(afinn, by = "word")



```

## 6. Sum Sentiment Per Sentence

```{r}
# Compute sentence-level sentiment scores
sentence_scores <- scored_words %>%
  group_by(sentence_id) %>%
  summarise(sentiment = sum(value), .groups = "drop") %>%
  left_join(words %>% distinct(sentence_id, student_id), by = "sentence_id") %>%
  left_join(reflections, by = "student_id") %>%
  select(student_id, question, sentence_id, sentiment, reflection)


```


## 7. Top Positive Sentences (AFINN)
```{r}
# Display top 5 most positive sentences
top_sentences <- sentence_scores %>%
  arrange(desc(sentiment)) %>%
  distinct(sentence_id, .keep_all = TRUE) %>%
  slice_head(n = 5)

top_sentences


```

## 8. Histogram of AFINN Sentiment Scores
```{r}

# Plot distribution of sentiment scores per sentence
ggplot(sentence_scores, aes(x = sentiment)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(
    title = "AFINN Sentiment Score Distribution",
    x = "Sentiment Score per Sentence",
    y = "Number of Sentences"
  )


```

## 9. NRC Emotion Analysis using 'syuzhet'

```{r}

# Apply NRC emotion scoring to entire reflections
nrc_emotions <- get_nrc_sentiment(reflections$reflection)

# Bind question and student_id for reference
nrc_df <- bind_cols(reflections %>% select(student_id, question), nrc_emotions)

# Preview the result
head(nrc_df)



```


## 10. Emotion Summary Plot (NRC)

```{r}

# Summarize total emotions across all reflections
nrc_summary <- nrc_emotions %>%
  summarise(across(everything(), sum)) %>%
  pivot_longer(cols = everything(), names_to = "emotion", values_to = "count")

# Plot NRC emotion categories
ggplot(nrc_summary, aes(x = reorder(emotion, -count), y = count)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(
    title = "Emotion Counts in Student Reflections (NRC)",
    x = "Emotion",
    y = "Frequency"
  ) +
  theme_minimal()



```